Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
ispell

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{rel-time-church.ps}
\end{center}
\caption{Relative run-time of
optimized analyers to baseline for Church numeral example.}
\label{fig:churchtime}
\end{figure}

@@

In section~\ref{sec:eval} we evaluate this implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

Compared with the standard machine semantics, this definition is
different in the following ways:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,
\item continuations are heap-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicity parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
\end{itemize}

@@

\paragraph{Abstract interpretation} can be characterized by setting the implicit
parameters just as above, but dropping the $\maddr \not\in \msto$
condition.  This family of interpereters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when mutiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract interpretation.  In particular,
concrete intepretation is a kind of abstract interpretation.  So is an
interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
intepretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
For the duration of this paper, $\sqcup$ is a pointwise lifting of $\cup$:
\begin{align*}
\msto\sqcup \varnothing &= \msto \\
\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{ 
  \begin{array}{ll}
    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
  \end{array}\right.
\end{align*}

@@

Each step in the fixpointer for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterwards update the
store, we end up with much less allocation and computational overhead.

We represent changes as a list of pairs of address to set of storable
values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
each step. Applying the changes is straightforward:

@@

The reduction relation does have to change to accomodate this
technique, but only in this regard. The whole store is still given so
that lookups are possible. Example rule:

@@

We also lift $\machstep$ differently to accomodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

All the above techniques can be combined into a single implementation
of the small step reduction relation and simple variations of a
general fixpoint combinator. We implementated all of these ideas in an
analysis of core Scheme (call-by-value lambda calculus extended with
variable-arity functions, call/cc, vectors, conses, recursive binding,
various primitives) in just 3000 lines of Racket. Most lines are spent
defining the primitives and expected library functions used in our
benchmarks. Here we describe the unifying vocabulary for writing such
a framework, and some design decisions related to scaling to a
language with compound data.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate the values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of constraints (either inclusion or equality constraints).  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for contraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{rel-time-church.ps}
\end{center}
\caption{Relative run-time of
optimized analyzers to baseline for Church numeral example.}
\label{fig:churchtime}
\end{figure}

@@

In section~\ref{sec:eval} we evaluate this implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyzer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

Compared with the standard machine semantics, this definition is
different in the following ways:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,
\item continuations are heap-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
\end{itemize}

@@

\paragraph{Abstract interpretation} can be characterized by setting the implicit
parameters just as above, but dropping the $\maddr \not\in \msto$
condition.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract interpretation.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is an
interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
For the duration of this paper, $\sqcup$ is a point-wise lifting of $\cup$:
\begin{align*}
\msto\sqcup \varnothing &= \msto \\
\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{ 
  \begin{array}{ll}
    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
  \end{array}\right.
\end{align*}

@@

Each step in the fixpointer for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterward update the
store, we end up with much less allocation and computational overhead.

We represent changes as a list of pairs of address to set of storeable
values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
each step. Applying the changes is straightforward:

@@

The reduction relation does have to change to accommodate this
technique, but only in this regard. The whole store is still given so
that lookups are possible. Example rule:

@@

We also lift $\machstep$ differently to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

All the above techniques can be combined into a single implementation
of the small step reduction relation and simple variations of a
general fixpoint combinator. We implemented all of these ideas in an
analysis of core Scheme (call-by-value lambda calculus extended with
variable-arity functions, call/cc, vectors, conses, recursive binding,
various primitives) in just 3000 lines of Racket. Most lines are spent
defining the primitives and expected library functions used in our
benchmarks. Here we describe the unifying vocabulary for writing such
a framework, and some design decisions related to scaling to a
language with compound data.

@@

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate the values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of constraints (either inclusion or equality constraints).  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for constraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.
